---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rcentury

<!-- badges: start -->
<!-- badges: end -->

The goal of Rcentury is to run CENTURY 4.7 (version Fortran) from within R. That includes preparing treatments, schedule, weather and site characteristic files with **.100** extension, and then reading and loading the resulting **lis** file into your R session as a **data.frame**.

## Installation

The package is available at github for download.

```{r install}
devtools::install_github("https://github.com/emf-creaf/Rcentury.git")
```


```{r load}
library(Rcentury)
## basic example code
```

## Example files

## How to prepare a simulation

There are two main ways to prepare the files that are required to run a simulation successfully:

1. From scratch:
2. Using example files:



## Example

Once installed, you can load it as:


