---
title: "Example model uncertainty and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example uncertainty propagation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup_Rcentury}
library(Rcentury)
```



### Introduction

We can use the Rcentury package to assess how uncertainties (aka errors) in a model's parameters contribute to the overall uncertainty of its output. Mathematically, it should be possible to derive approximate analytical expressions for n-error error propagation by using Taylor expansions, which involves calculating analytical n-order derivatives. However, this approach quickly becomes cumbersome and impractical when the number of parameters grows high and, especially, when the function to evaluate is very complex and non-linear like CENTURY. Therefore, to determine error propagation for the CENTURY model we turn to a much simpler Monte Carlo numerical approach, which consists of feeding the model with input parameters of interests which vary randomly at each model run following a given distribution of known mean and variance.

For the sake of simplicity we will demonstrate how to perform an uncertainty propagation analysis of uncertainty in soil texture (i.e. clay, silt and sand) only, keeping all the other parameters of the model fixed. In this case, at each model run we will draw random proportion values for each of the three parameters, calculate the CENTURY model with those values and then save the output. This will be done many times to derive reliable confidence intervals for the outputs.


### How to draw multivariate proportions

Texture is always given as a set of 3 proportions which add up to 1 always, we need to specify a random distribution that accounts for that property. Consequently, we turn to Dirichlet distributions, which 




The means of the proportions will be fixed in our simulations.

$$f(x_1, \dots, x_K; \alpha_1, \dots, \alpha_K) = \frac{1}{B(\boldsymbol{\alpha})} \prod_{i=1}^K x_i^{\alpha_i - 1}$$

where:

$$B(\boldsymbol{\alpha}) = \frac{\prod_{i=1}^K \Gamma(\alpha_i)}{\Gamma(\sum_{i=1}^K \alpha_i)}$$

and:

$$x_i \in [0, 1], \quad \text{for all } i \in \{1, \dots, K\}$$

as well as:

$$\sum_{i=1}^K x_i = 1$$

The mean of the $x_i$ is, then:

$$E[x_i] = \frac{\alpha_i}{\sum_{j=1}^K \alpha_j}$$
and the expression for the variance is:

$$Var[x_i] = \frac{\tilde{\alpha_i}\cdot(1-\tilde{\alpha_i})}{\alpha_0-1}$$

where:

$$\tilde{\alpha_i}=\frac{\alpha_i}{\alpha_0}$$
We can use those two expressions to calculate the $\alpha_j$ values when we are given four parameters for the $E[x_i]$ or $E[V_i]$. The Rcentury package provides with a function, called $dirichlet\_params.R$, that can calculate the $\alpha_j$ values when all the means $E[x_i]\text{ for all } i \in \{1, \dots, K\}$ and one single variance value $Var[x_j]$ are known.


The Dirichlet distribution is implemented in package extraDistr, which we can install from CRAN:

```{r install_distrib}
# install.packages("extraDistr")
library(extraDistr)
```

For example, for $K=3$ and $E[x_1]=.35$, $E[x_2]=.12$, $E[x_3]=.53$ and $Var[x_2]=.17$, we can easily generate sequences of Dirichlet-distributed triplets.

```{r hist_triplets, fig.width = 7}
alpha <- dirichlet_params(c(.35, .12, .53), c(NA, .042, NA))
x <- rdirichlet(10000, alpha)
par(mfrow = c(1, 3))
hist(x[, 1], xlab = "", main = "First component")
hist(x[, 2], xlab = "", main = "Second component")
hist(x[, 3], xlab = "", main = "Third component")
par(mfrow = c(1,1))
```


### Rcentury simulations



```{r copy_files}
from_dir <- system.file("extdata/4.forest",  package = "Rcentury")
to_dir <- tempdir()
dir.create(to_dir)
invisible(file.copy(from_dir, to_dir, overwrite = TRUE, recursive = TRUE))
path_exe <- system.file("bin/windows",  package = "Rcentury")
century_exe <- file.path(path_exe, "century_47.exe")
list100_exe <- file.path(path_exe, "list100_47.exe")
to_dir <- file.path(to_dir, "4.forest")
invisible(file.copy(century_exe, to_dir, overwrite = TRUE, recursive = TRUE))
invisible(file.copy(list100_exe, to_dir, overwrite = TRUE, recursive = TRUE))
```



```{r simu}
#

nsimu <- 10
for (i in 1:nsimu) {
  df <- century_run(to_dir, "duke.sch", "res.bin", "res.lis", "outvars.txt", verbose = FALSE)
  invisible(file.remove(file.path(to_dir, c("res.bin", "res.lis"))))
}
```
